### 2017-08-16 body.tex
Remove acknowledgement of NSF of China :O

\section{Introduction}
\subsection{Motivation and Background}
Since the Fortran 2008 standard was published in 2010~\cite{iso2010fortran}, a handful of studies in the open literature
have provided encouraging assessments of Fortran's intrinsic parallel programming model, \gls{caf}, running in research
applications running at scale~\cite{preissl2011multithreaded,garain2015comparing,mozdzynski2015partitioned}.   One of the
most attractive features of \gls{caf} is the ability to express parallel algorithms once within the confines of a single
language standard without embedding the use of compiler directives or communication layers defined outside the Fortran
Fortran language standard (e.g., no \gls{mpi} or OpenMP.  In theory, this should leave the choice of communication layers
as a link-time decision.  Ideally, an application developer writes a standard Fortran-only application, compiles it with any
standard-conforming compiler, and possibly reserves the option to incorporate any one (or more) of several communication layers
at link-time.

Such flexibility poses the computing equivalent of what food author Michael Pollen termed ``The Omnivore's Dilemma:'' if I
belong to a species that owes some of its evolutionary advantage to being able to eat a wide variety of foods, what food is
best for me to eat?  Similarly, if I can express my parallel algorithm once using \gls{caf} and then consume cycles atop a
multitude of software stacks and hardware platforms, where best to consume cycles and using which supporting software stack?
Here we report the results of an initial study of the current options for compiling, linking, and executing a \gls{mini-app}
designed to be representative of the parallel numerical algorithms and physics employed in the \gls{icar} being developed at
the \gls{ncar}.

\gls{icar} simulates the motion of the atmosphere at kilometer length scales and produces flow patterns with a fidelity that is
attractive to the hydrology community, which studies surface water.  The animation in Figure~\ref{figure:icar} depicts atmospheric processes over North America.

\begin{figure}
   \vspace{-18pt}
   % Animate frames at 7 fps created via 'ffmpeg -i qvmovie.mp4 -r 7 frame%d.pngr', add control buttons,
   % and automatically play when the page is viewed:
   \vbox{\hspace{-24pt}
   %\animategraphics[width=1.1\columnwidth,controls,autoplay]{7}{figures/icar/frames/frame}{1}{30}
   \includegraphics[width=1.1\columnwidth]{figures/icar/frames/frame22.png}
   }
   \caption{An animation of atmospheric flow of water vapor (blues) and resulting precipitation (green to red) simulated by the \href{https://github.com/gutmann/icar}{\gls{icar}}. {\it For the animated version of this graphic, please access https://github.com/sourceryinstitute/coarray-icar-paw17/releases/download/1.0/main.pdf and use Adobe Acrobat Reader 7.0 or later.}
\label{figure:icar}}
\end{figure}

\begin{figure*}
  \includegraphics[width=\textwidth]{figures/fig1_speedup.png}
  \includegraphics[width=\textwidth]{figures/fig2_timing.png}
  \caption{Speedup and timing results for two different domain sizes, two different communications backends (OSH=OpenSHMEM, MPI=MPI) and two different communication methods (``get'' and ``put'') using \href{https://github.com/gutmann/coarray_icar}{Coarray ICAR}.\label{fig1-2}}
\end{figure*}

\begin{figure*}
  \includegraphics[width=\textwidth]{figures/fig3_cross_platform.png}
  \includegraphics[width=\textwidth]{figures/fig4_extreme_scaling.png}
   \caption{Cross-platform and extreme scaling results using \href{https://github.com/gutmann/coarray_icar}{Coarray ICAR}.\label{fig3-4}}
\end{figure*}

\subsection{Objectives}

\section{Methodology}
\subsection{Numerical algorithms}
The core numerics of \gls{icar} implemented here are the Thompson Eidhammer microphysics parameterization~\cite{Thompson:2014cw} and the first order MP-DATA advection algorithm~\cite{Smolarkiewicz:1998il}.
The microphysics parameterization requires 11 primary input variables (e.g.\ air pressure, water vapor, cloud water), 9 of which are modified by the parameterization and advected throughout the domain requiring communication between processes.

The test case implemented here uses a classical idealized hill case.
An idealized 1000 m high mountain is generated in the middle of the domain defined by a sine curve.
The air is initialized to near 100\% relative humidity, and a background wind of approximately 10 m s$^{-1}$ is specified.

At every time step, the outer edges of the local domain are processed first,
then passed to their neighbors while the interior of the domain is processed.
This is tested using both a \gls{caf} ``put'' operation, in which one-sided communication is initialized asynchronously to send data to a neighbor,
and a \gls{caf} ``get'' operation, in which one-sided blocking communication is used to retrieve data from a neighbor.

\subsection{Compilers, runtimes, and hardware}

We performed the experiments presented here on two supercomputers at at the National Energy Research Scientific Computing Center (NERSC),
located at Lawrence Berkeley National Laboratory and one supercomputer at the National Center for Atmospheric Research (NCAR).
One system is ``Edison,'' a Cray XC30 featuring 5586 nodes with two sockets of 12-core Intel Xeon Processor E5-2695 v2 (``Ivy Bridge''), running at \num{2.4}~\si{\giga\hertz}.
The second system is ``Cori,'' a Cray XC40 which contains \num{12076} compute nodes, spanning two architectures. \num{2388} nodes each have two sockets of 16-core Intel Xeon Processor E5-2698 v3 (``Haswell'') operating at \num{2.3}~\si{\giga\hertz}; the other \num{9688} nodes are single-socket, 68-core Intel Xeon Phi Processor 7250 (``Knights Landing'') at \num{1.4}~\si{\giga\hertz}.
Both Edison and Cori feature the same ``Aries'' interconnect, with a dragonfly topology.
All measurements reported here which ran on Cori were executed on the Xeon Phi nodes, running in the ``quadrant`` NUMA configuration, with the MCDRAM configured as a transparent cache to the DDR4 memory.
The third system is ``Cheyenne,'' an SGI ICE XA Cluster with \num{4032} compute nodes, with two sockets of 18-core Intel Xeon Processor E5-2697 v4 (``Broadwell''), running at \num{2.3}~\si{\giga\hertz}.
Cheyenne uses a Mellanox EDR infiniband interconnect with a partial 9D Enhanced Hypercube single-plane topology.
We compiled the ICAR code on both systems at NERSC using version 8.6.0 of the Cray Fortran compiler, and at NCAR using version 6.3 of the gfortran compiler.
gfortran uses the opencoarrays library to provide coarray fortran support, and we tested the code using both SGI MPT MPI and OpenSHMEM implementations.
In all runs, we ran the code with one image per physical core (68 images per Xeon Phi node on Cori, 24 images per Xeon node on Edison, 36 images per node on Cheyenne).

\section{Discussion of Results}
Results with two different domain sizes (500 x 500 and 2000 x 2000 grid cells) are presented in figure~\ref{fig1-2} for simulations performed on Cheyenne.
These tests compare scaling with blocking ``get'' and non-blocking ``put'' operations, and compare scaling using MPI or OpenSHMEM for communication.
The differences between MPI and OpenSHMEM are minor, with OpenSHMEM performing only slightly better.
However, the differences between the ``get'' and ``put'' operations are substantial.
For the 500x500 domain, ``puts'' scale well to 1800 cores, then scale poorly to \num{3600} cores;
however, ``gets'' only scale well to \num{504} cores, scale poorly to \num{1224}, an no further improvement comes from up to \num{3600} cores.
Note MPI ``gets'' are not reported because a problem with the MPI implementation makes them several orders of magnitude less efficient,
similarly, MPI ``puts'' results are not reported beyond \num{1800} cores because a problem in the MPI implementation caused memory allocations to become a bottleneck at higher process counts.
For the 2000x2000 domain, the same pattern is observed but scaling of ``puts'' remains strong up to 3600 cores;
scaling of ``gets'' continuous, albeit poorly, to \num{3600} cores.

Results across multiple compilers, machines, and architectures are presented in figure~\ref{fig3-4} (top).
In this case, only ``puts'' are tested, and only for the 2000x2000 domain.
The GNU compiler was used on Cheyenne (Xeons), and the Cray compiler was used on both Edison (Xeons) and Cori (KNL).
To aid interpretation, the scaling results are reported as a fraction of the ideal scaling for each machine in the top left.
These results show that the Cray compiler+system scales better than the gfortran+SGI system,
with 75\% efficiency at >10k cores, while on Cheyenne, only 55\% of ideal was achieved.
This also shows that the KNL system scales well out to large core counts (60\% ideal with \num{19200} cores),
but that the total runtimes are significantly slower then the equivalent runtimes on Xeons (top right).
The KNL performance may be improved in the future by implementing OpenMP threaded parallelism within a node.

To further test the scaling performance in the best case, ``puts'' on Cray Xeons, a limited set of tests were performed up to nearly \num{100000} cores (fig.~\ref{fig3-4} bottom).
This system scaled well to over \num{10000} cores, and continued scaling with nearly 50\% of single node efficiency obtained at \num{98304} cores.

\section{Conclusions}
We have presented scaling and performance analysis using \gls{caf} on different architectures, compilers, and communications backends.
These tests have utilized modern Fortran standards and as a result no changes to the code were required for different configurations.

This work demonstrates the capability of one-sided non-blocking communications protocols to readily scale up to very high core counts.
It is expected that with moderate performance optimizations and a larger total problem size, this system will be able scale to even higher core counts.

%%%
%%% Sample tables
%%%

%\begin{table}
%  \caption{Frequency of Special Characters}
%  \label{tab:freq}
%  \begin{tabular}{ccl}
%    \toprule
%    Non-English or Math&Frequency&Comments\\
%    \midrule
%    \O & 1 in 1,000& For Swedish names\\
%    $\pi$ & 1 in 5& Common in math\\
%    \$ & 4 in 5 & Used in business\\
%    $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
%  \bottomrule
%\end{tabular}
%\end{table}

%\begin{table*}
%  \caption{Some Typical Commands}
%  \label{tab:commands}
%  \begin{tabular}{ccl}
%    \toprule
%    Command &A Number & Comments\\
%    \midrule
%    \texttt{{\char'134}author} & 100& Author \\
%    \texttt{{\char'134}table}& 300 & For tables\\
%    \texttt{{\char'134}table*}& 400& For wider tables\\
%    \bottomrule
%  \end{tabular}
%\end{table*}
% end the environment with {table*}, NOTE not {table}!

%It is strongly recommended to use the package booktabs~\cite{Fear05}
%and follow its main principles of typography with respect to tables:
%\begin{enumerate}
%\item Never, ever use vertical rules.
%\item Never use double rules.
%\end{enumerate}
%It is also a good idea not to overuse horizontal rules.

%%%
%%% Sample figures
%%%

%\begin{figure}
%\includegraphics{fly}
%\caption{A sample black and white graphic.}
%\end{figure}

%\begin{figure}
%\includegraphics[height=1in, width=1in]{fly}
%\caption{A sample black and white graphic
%that has been resized with the \texttt{includegraphics} command.}
%\end{figure}

%\begin{figure*}
%\includegraphics{flies}
%\caption{A sample black and white graphic
%that needs to span two columns of text.}
%\end{figure*}
%
%\begin{figure}
%\includegraphics[height=1in, width=1in]{rosette}
%\caption{A sample black and white graphic that has
%been resized with the \texttt{includegraphics} command.}
%\end{figure}

%\end{document}  % This is where a 'short' article might terminate

%\appendix
%Appendix A
%\section{Code snippets: collective subroutines?}
% This next section command marks the start of
% Appendix B, and does not continue the present hierarchy
%\section{Anything else?}

\begin{acks}
  The first author would like to thank the Visitor Programs of the Computational Information Systems Laboratory
  and the Research Applications Laboratory of \gls{ncar} for travel support for a visit during which much of
  the work presented in this paper was performed.

  This work as funded in part through a contract from the U.S. Army Corps of Engineers.

\end{acks}

### 2017-08-16 main.tex
Delete extraneous word in abstract

\documentclass[sigconf, authordraft]{acmart}

\usepackage{booktabs} % For formal tables
\usepackage{animate}
\usepackage{siunitx} % for SI formatting of digits and units

\usepackage[acronym,nonumberlist]{glossaries}
\makeglossaries

\input glossary/glossary.tex
\input glossary/acronyms.tex

% Copyright
%\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

% DOI
\acmDOI{xx.xxx/xxx_x}

% ISBN
\acmISBN{xxx-xxxx-xx-xxx/xx/xx}

%Conference
\acmConference[PAW17]{PGAS Applications Workshop}{November 2017}{Denver, Colorado USA}
\acmYear{2017}
\copyrightyear{2017}

%\acmPrice{15.00}

\acmSubmissionID{xxx-xxx-xx}

\begin{document}
\title{Performance portability of an intermediate-complexity atmospheric research model in coarray Fortran}
%\titlenote{Produces the permission block, and
%  copyright information}
\subtitle{Extended Abstract}
%\subtitlenote{The full version of the author's guide is available as
%  \texttt{acmart.pdf} document}

\author{Damian Rouson}
\orcid{0000-0002-2234-868X}
\affiliation{%
  \institution{Sourcery Institute}
  \streetaddress{2323 Broadway}
  \city{Oakland}
  \state{California}
  \postcode{94612}
}
\email{damian@sourceryinstitute.org}
\renewcommand{\shortauthors}{D. Rouson et al.}

\author{Ethan D Gutmann}
\orcid{0000-0003-4077-3430}
\affiliation{%
  \institution{National Center for Atmospheric Research}
  \streetaddress{3450 Mitchell Ln.}
  \city{Boulder}
  \state{Colorado}
  \postcode{80301}
}
\email{gutmann@ucar.edu}

\author{Brian Friesen}
\orcid{0000-0002-1572-1631}
\affiliation{%
  \institution{Lawrence Berkeley National Laboratory}
  \streetaddress{1 Cyclotron Road, Mail Stop 59R4010A}
  \city{Berkeley}
  \state{California}
  \postcode{94720}
}
\email{BFriesen@lbl.gov}

\author{Alessandro Fanfarillo}
\orcid{0000-0003-3487-7452}
\affiliation{%
  \institution{National Center for Atmospheric Research}
  \streetaddress{ 1850 Table Mesa Dr.}
  \city{Boulder}
  \state{Colorado}
  \postcode{80305}
}
\email{elfanfa@ucar.edu}

\begin{abstract}
We present results on the scalability and performance of an open-source, \gls{caf} \gls{mini-app} that solves
several parallel, numerical algorithms known to dominate the execution of the \gls{icar}~\cite{gutmann2016intermediate}.
The solver employs standard Fortran 2008 features and includes several Fortran 2008 implementations of the collective
subroutines defined in the Committee Draft of the upcoming Fortran 2015 standard.  The ability of \gls{caf} to run atop various
communication layers and the increasing compiler support for \gls{caf} facilitated initial evaluations of several compliers,
runtime libraries and hardware platforms.  Results are presented for the GNU and Cray compilers, each of which offers
different parallel runtime libraries employing one or more communication layers, including \gls{mpi}, OpenSHMEM, and proprietary
alternatives.  We studied performance on both multi- and many-core processors running on distributed-memory systems.  The
results of initial simulations suggest promising scaling behavior across a range of hardware, compiler, and runtime choices
on platforms ranging up to 100,000 cores.
\end{abstract}

%
% The code below was generated by the tool at
% http://dl.acm.org/ccs.cfm
%
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008.10011009.10010175</concept_id>
<concept_desc>Software and its engineering~Parallel programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10010405.10010432.10010437.10010438</concept_id>
<concept_desc>Applied computing~Environmental sciences</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Parallel programming languages}
\ccsdesc[300]{Applied computing~Environmental sciences}

\keywords{coarray Fortran, computational hydrometeorology, parallel programming}

\maketitle

\input{body}

\bibliographystyle{ACM-Reference-Format}
\bibliography{bibliography}

\end{document}

### 2017-08-16 main.tex
update abstract

@@
 \begin{abstract}
-We present results on the scalability and performance of an open-source, \gls{caf} \gls{mini-app} that solves
-several parallel, numerical algorithms known to dominate the execution of the \gls{icar}~\cite{gutmann2016intermediate}.
-The solver employs standard Fortran 2008 features and includes several Fortran 2008 implementations of the collective
+We examine the scalability and performance of an open-source, \gls{caf} \gls{mini-app} that solves
+several parallel, numerical algorithms known to dominate the execution of the \gls{icar}~\cite{gutmann2016intermediate} model,
+a package developed at the \gls{ncar}.
+The \gls{mini-app} employs standard Fortran 2008 features and includes several Fortran 2008 implementations of the collective
 subroutines defined in the Committee Draft of the upcoming Fortran 2015 standard.  The ability of \gls{caf} to run atop various
-communication layers and the increasing compiler support for \gls{caf} facilitated initial evaluations of several compliers,
+communication layers and the increasing availability of \gls{caf} compilers facilitated evaluating several compilers,
 runtime libraries and hardware platforms.  Results are presented for the GNU and Cray compilers, each of which offers
 different parallel runtime libraries employing one or more communication layers, including \gls{mpi}, OpenSHMEM, and proprietary
-alternatives.  We studied performance on both multi- and many-core processors running on distributed-memory systems.  The
-results of initial simulations suggest promising scaling behavior across a range of hardware, compiler, and runtime choices
+alternatives.  We studied performance on multi- and many-core processors in distributed-memory systems.  The
+results show promising scaling behavior across a range of hardware, compiler, and runtime choices
 on platforms ranging up to 100,000 cores.

### 2017-08-16 main.tex
Update abstract

@@
 % DOI
-\acmDOI{xx.xxx/xxx_x}
+%\acmDOI{xx.xxx/xxx_x}
 
 % ISBN
-\acmISBN{xxx-xxxx-xx-xxx/xx/xx}
+%\acmISBN{xxx-xxxx-xx-xxx/xx/xx}
 
@@
 a package developed at the \gls{ncar}.
-The \gls{mini-app} employs standard Fortran 2008 features and includes several Fortran 2008 implementations of the collective
+The \gls{mini-app} uses standard Fortran 2008, including several Fortran 2008 implementations of the collective
 subroutines defined in the Committee Draft of the upcoming Fortran 2015 standard.  The ability of \gls{caf} to run atop various
-communication layers and the increasing availability of \gls{caf} compilers facilitated evaluating several compilers,
+communication layers and the increasing \gls{caf} compiler availability facilitated evaluating several compilers,
 runtime libraries and hardware platforms.  Results are presented for the GNU and Cray compilers, each of which offers
 different parallel runtime libraries employing one or more communication layers, including \gls{mpi}, OpenSHMEM, and proprietary
-alternatives.  We studied performance on multi- and many-core processors in distributed-memory systems.  The
+alternatives.  We studied performance on multi- and many-core processors in distributed memory.  The
 results show promising scaling behavior across a range of hardware, compiler, and runtime choices
@@
 
-\keywords{coarray Fortran, computational hydrometeorology, parallel programming}
+\keywords{coarray Fortran, computational hydrometeorology}
 

### 2017-08-16 body.tex
Revise intro & add omnivore's dilemma citation

@@
 \subsection{Motivation and Background}
-Since the Fortran 2008 standard was published in 2010~\cite{iso2010fortran}, a handful of studies in the open literature
-have provided encouraging assessments of Fortran's intrinsic parallel programming model, \gls{caf}, running in research
+Since the Fortran 2008 standard was published in 2010~\cite{iso2010information}, a handful of studies in the open literature
+have provided encouraging assessments of Fortran's intrinsic parallel programming model, \gls{caf}, in research
 applications running at scale~\cite{preissl2011multithreaded,garain2015comparing,mozdzynski2015partitioned}.   One of the
 most attractive features of \gls{caf} is the ability to express parallel algorithms once within the confines of a single
-language standard without embedding the use of compiler directives or communication layers defined outside the Fortran
-Fortran language standard (e.g., no \gls{mpi} or OpenMP.  In theory, this should leave the choice of communication layers
-as a link-time decision.  Ideally, an application developer writes a standard Fortran-only application, compiles it with any
+language standard without embedding compiler directives or referencing communication layers that are not part of the
+language standard (e.g., no \gls{mpi} or OpenMP).  In theory, this should leave the choice of communication layers
+as a link-time decision.  Ideally, an application developer writes a standard, Fortran-only application, compiles it with any
 standard-conforming compiler, and possibly reserves the option to incorporate any one (or more) of several communication layers
@@
 
-Such flexibility poses the computing equivalent of what food author Michael Pollen termed ``The Omnivore's Dilemma:'' if I
+Such flexibility poses the computing equivalent of what food author Michael Pollan termed ``The Omnivore's Dilemma:'' if I
 belong to a species that owes some of its evolutionary advantage to being able to eat a wide variety of foods, what food is
-best for me to eat?  Similarly, if I can express my parallel algorithm once using \gls{caf} and then consume cycles atop a
+best for me to eat?~\cite{pollan2006omnivore}  Similarly, if I can express my parallel algorithm once using \gls{caf} and then consume cycles atop a
 multitude of software stacks and hardware platforms, where best to consume cycles and using which supporting software stack?
-Here we report the results of an initial study of the current options for compiling, linking, and executing a \gls{mini-app}
-designed to be representative of the parallel numerical algorithms and physics employed in the \gls{icar} being developed at
-the \gls{ncar}.
+Here we report the results of an initial study of several current options for compiling, linking, and executing a \gls{mini-app}
+designed to be representative of the parallel numerical algorithms and physics employed in \gls{icar}.
 
 \gls{icar} simulates the motion of the atmosphere at kilometer length scales and produces flow patterns with a fidelity that is
-attractive to the hydrology community, which studies surface water.  The animation in Figure~\ref{figure:icar} depicts atmospheric processes over North America.
+attractive to the hydrology community, which studies surface water.  The animation in Figure~\ref{figure:icar} depicts atmospheric
+processes over North America as simulated by \gls{icar}.
 

### 2017-08-16 glossary/acronyms.tex
Revise intro & add omnivore's dilemma citation

\newacronym{caf}{CAF}{coarray Fortran}
\newacronym{mini-app}{mini-app}{mini-application}
\newacronym{ncar}{NCAR}{National Center for Atmospheric Research}
\newacronym{wrf-hydro}{WRF-Hydro}{Weather Research and Forcecasting -- Hydrological Model}
\newacronym{spmd}{SPMD}{Single-Program Multiple-Data}
\newacronym{url}{URL}{Uniform Resource Locator}
\newacronym{tdd}{TDD}{test-driven development}
\newacronym{upc++}{UPC++}{Unified Parallel C++}
\newacronym{upc}{UPC}{Unified Parallel C}
\newacronym{lbnl}{LBNL}{Lawrence Berkeley National Laboratoory}
\newacronym{ide}{IDE}{integrated development environment}
\newacronym{html}{HTML}{Hypertext Markup Language}
\newacronym{icar}{ICAR}{The Intermediate Complexity Atmospheric Research}
\newacronym{exaflops}{exaFLOPS}{$10^{18}$ floating-point operations per second}
\newacronym{hpc}{HPC}{high-performance computing}
\newacronym{nasa}{NASA}{National Aeronautics and Space Administration}
\newacronym{nersc}{NERSC}{National Energy Research Scientific Computing Center}
\newacronym{doe}{DOE}{Department of Energy}
\newacronym{us}{US}{United States}
\newacronym{oop}{OOP}{Object-Oriented Programming}
\newacronym{ood}{OOD}{Object-Oriented Design}
\newacronym{uml}{UML}{Unified Modeling Language}
\newacronym{psblas}{PSBLAS}{Parallel Sparse Basic Linear Algebra Subroutines}
\newacronym{gpu}{GPU}{Graphics Processing Unit}
\newacronym{gpgpu}{GPGPU}{General-Purpose Graphics Processing Unit}
\newacronym{simt}{SIMT}{Single-Instruction, Multiple-Threads}
\newacronym{simd}{SIMD}{Single-Instruction, Multiple-Data}
\newacronym{2d}{2D}{Two-Dimensional}
\newacronym{ram}{RAM}{Random-Access Memory}
\newacronym{cpu}{CPU}{Central Processing Unit}
\newacronym{adsm}{ADSM}{Asymmetric Distributed Shared Memory}
\newacronym{loc}{LOC}{Lines Of Code}
\newacronym{csr}{CSR}{Compressed Storage by Rows}
\newacronym{coo}{COO}{COOrdinate}
\newacronym{csc}{CSC}{Compressed Sparse Columns}
\newacronym{ell}{ELL}{ELLpack}
\newacronym{jad}{JAD}{JAgged Diagonals}
\newacronym{cuda}{CUDA}{Compute Unified Device Architecture}
\newacronym{gflops}{GFLOPS}{Giga (billions) of Floating-point Operations Per Second, or GigaFLOPS}
\newacronym{pde}{PDE}{Partial Differential Equation}
\newacronym{spmv}{SpMV}{Sparse Matrix-Vector multiplication}
\newacronym{ecc}{ECC}{Error-Correcting Code}
\newacronym{petsc}{PETSc}{the Portable Extensible Toolkit for Scientific Computation}
\newacronym{flame}{FLAME}{the Formal Linear Algebra Methods Environment}
\newacronym{cfd}{CFD}{computational fluid dynamics}

### 2017-08-16 body.tex
Add Sec 1.2 Objectives; edit 2.1 Physics &Numerics

@@
 \subsection{Motivation and Background}
-Since the Fortran 2008 standard was published in 2010~\cite{iso2010information}, a handful of studies in the open literature
+Since the Fortran 2008 standard was published in 2010~\cite{iso2010information}, a handful of published studies
 have provided encouraging assessments of Fortran's intrinsic parallel programming model, \gls{caf}, in research
@@
 \subsection{Objectives}
+The goals of this work are to investigate and evaluate several software and hardware alternatives for compiling and linking the
+same \gls{mini-app} \gls{caf} source code using the following technologies:
+\begin{itemize}
+  \item \gls{mpi}~\cite{mpiforum2016mpi} and OpenSHMEM~\cite{chapman2010introducing} communication layers,
+  \item \gls{gcc} and Cray compilers, and
+  \item Multi- and many-core processors.
+\end{itemize}
+Additionally we show performance differences between two one-sided data access patterns:
+\begin{enumerate}
+  \item Gets: an image that needs data proactively gets the data from memory managed by a providing
+        image without the providing image's active involvement, and
+  \item Puts: an image that is able to provide data proactively puts the data into memory managed by
+        a receiving image without the receiving image's active involvement.
+\end{enumerate}
+This study is a work-in-progress that will guide future decisions around the choice of compiler, communication
+layer, platform, and data access patterns.
 
 \section{Methodology}
-\subsection{Numerical algorithms}
-The core numerics of \gls{icar} implemented here are the Thompson Eidhammer microphysics parameterization~\cite{Thompson:2014cw} and the first order MP-DATA advection algorithm~\cite{Smolarkiewicz:1998il}.
+\subsection{Physics and numerics}
+The coarray-\gls{icar} implemention employed here uses the Thompson Eidhammer microphysics parameterization~\cite{Thompson:2014cw} and the first-order MP-DATA advection algorithm~\cite{Smolarkiewicz:1998il}.
 The microphysics parameterization requires 11 primary input variables (e.g.\ air pressure, water vapor, cloud water), 9 of which are modified by the parameterization and advected throughout the domain requiring communication between processes.
 
-The test case implemented here uses a classical idealized hill case.
+The test case implemented here uses a classical, idealized hill.
 An idealized 1000 m high mountain is generated in the middle of the domain defined by a sine curve.

### 2017-08-16 body.tex
Edit Section 2.2 Compilers, runtimes, and hardware

@@
 \subsection{Objectives}
-The goals of this work are to investigate and evaluate several software and hardware alternatives for compiling and linking the
+The goal of this work is to evaluate several alteranatives for compiling, linking, and executing the
 same \gls{mini-app} \gls{caf} source code using the following technologies:
@@
 \end{enumerate}
-This study is a work-in-progress that will guide future decisions around the choice of compiler, communication
+Our results will inform our future decisions around the choice of compiler, communication
 layer, platform, and data access patterns.
@@
 
-We performed the experiments presented here on two supercomputers at at the National Energy Research Scientific Computing Center (NERSC),
-located at Lawrence Berkeley National Laboratory and one supercomputer at the National Center for Atmospheric Research (NCAR).
-One system is ``Edison,'' a Cray XC30 featuring 5586 nodes with two sockets of 12-core Intel Xeon Processor E5-2695 v2 (``Ivy Bridge''), running at \num{2.4}~\si{\giga\hertz}.
-The second system is ``Cori,'' a Cray XC40 which contains \num{12076} compute nodes, spanning two architectures.
+We performed the experiments presented here on two supercomputers at \gls{nersc},
+located at \gls{lbnl}, and one supercomputer at \gls{ncar}.
+One \gls{lbnl} system is ``Edison,'' a Cray XC30 featuring 5586 nodes with two sockets of 12-core Intel Xeon Processor E5-2695 v2 (``Ivy Bridge''), running at \num{2.4}~\si{\giga\hertz}.
+The second \gls{lbnl} system is ``Cori,'' a Cray XC40 which contains \num{12076} compute nodes, spanning two architectures.
 \num{2388} nodes each have two sockets of 16-core Intel Xeon Processor E5-2698 v3 (``Haswell'') operating at \num{2.3}~\si{\giga\hertz}; the other \num{9688} nodes are single-socket, 68-core Intel Xeon Phi Processor 7250 (``Knights Landing'') at \num{1.4}~\si{\giga\hertz}.
@@
 All measurements reported here which ran on Cori were executed on the Xeon Phi nodes, running in the ``quadrant`` NUMA configuration, with the MCDRAM configured as a transparent cache to the DDR4 memory.
-The third system is ``Cheyenne,'' an SGI ICE XA Cluster with \num{4032} compute nodes, with two sockets of 18-core Intel Xeon Processor E5-2697 v4 (``Broadwell''), running at \num{2.3}~\si{\giga\hertz}.
+
+The \gls{ncar} system is ``Cheyenne,'' an SGI ICE XA Cluster with \num{4032} compute nodes, with two sockets of 18-core Intel Xeon Processor E5-2697 v4 (``Broadwell''), running at \num{2.3}~\si{\giga\hertz}.
 Cheyenne uses a Mellanox EDR infiniband interconnect with a partial 9D Enhanced Hypercube single-plane topology.
-We compiled the ICAR code on both systems at NERSC using version 8.6.0 of the Cray Fortran compiler, and at NCAR using version 6.3 of the gfortran compiler.
-gfortran uses the opencoarrays library to provide coarray fortran support, and we tested the code using both SGI MPT MPI and OpenSHMEM implementations.
-In all runs, we ran the code with one image per physical core (68 images per Xeon Phi node on Cori, 24 images per Xeon node on Edison, 36 images per node on Cheyenne).
+We compiled coarray-\gls{icar} on the \gls{nersc} systems using the Cray Fortran compiler version 8.6.0.  We compiled
+at \gls{ncar} using of the \gls{gcc} Fortran (gfortran) compiler front end version 6.3, which uses the
+OpenCoarrays \gls{abi}~\cite{fanfarillo2014opencoarrays} for to support \gls{caf}.  We tested two OpenCoarrays parallel runtime libraries:
+\begin{enumerate}
+  \item The default \gls{mpi} library using the SGI MPT \gls{mpi},
+  \item The openshmem branch using OpenSHMEM.
+\end{enumerate}
+We distributed \gls{caf} images as follows:
+\begin{itemize}
+  \item 68 images per Xeon Phi node on Cori,
+  \item 24 images per Xeon node on Edison, and
+  \item 36 images per node on Cheyenne.
+\end{itemize}
+corresponding to one image per physical core in each case.
 
@@
 Results with two different domain sizes (500 x 500 and 2000 x 2000 grid cells) are presented in figure~\ref{fig1-2} for simulations performed on Cheyenne.
-These tests compare scaling with blocking ``get'' and non-blocking ``put'' operations, and compare scaling using MPI or OpenSHMEM for communication.
-The differences between MPI and OpenSHMEM are minor, with OpenSHMEM performing only slightly better.
+These tests compare scaling with blocking ``get'' and non-blocking ``put'' operations, and compare scaling using \gls{mpi} or OpenSHMEM for communication.
+The differences between \gls{mpi} and OpenSHMEM are minor, with OpenSHMEM performing only slightly better.
 However, the differences between the ``get'' and ``put'' operations are substantial.
@@
 however, ``gets'' only scale well to \num{504} cores, scale poorly to \num{1224}, an no further improvement comes from up to \num{3600} cores.
-Note MPI ``gets'' are not reported because a problem with the MPI implementation makes them several orders of magnitude less efficient,
-similarly, MPI ``puts'' results are not reported beyond \num{1800} cores because a problem in the MPI implementation caused memory allocations to become a bottleneck at higher process counts.
+Note \gls{mpi} ``gets'' are not reported because a problem with the \gls{mpi} implementation makes them several orders of magnitude less efficient,
+similarly, \gls{mpi} ``puts'' results are not reported beyond \num{1800} cores because a problem in the \gls{mpi} implementation caused memory allocations to become a bottleneck at higher process counts.
 For the 2000x2000 domain, the same pattern is observed but scaling of ``puts'' remains strong up to 3600 cores;

### 2017-08-16 main.tex
Edit Section 2.2 Compilers, runtimes, and hardware

@@
 We examine the scalability and performance of an open-source, \gls{caf} \gls{mini-app} that solves
-several parallel, numerical algorithms known to dominate the execution of the \gls{icar}~\cite{gutmann2016intermediate} model,
+several parallel, numerical algorithms known to dominate the execution of \gls{icar}~\cite{gutmann2016intermediate} model,
 a package developed at the \gls{ncar}.

### 2017-10-03 body.tex
Committing edits from August 16.

@@
    \vbox{\hspace{-24pt}
-   %\animategraphics[width=1.1\columnwidth,controls,autoplay]{7}{figures/icar/frames/frame}{1}{30}
+   %\animategraphics[width=\columnwidth,controls,autoplay]{7}{figures/icar/frames/frame}{1}{30}
    \includegraphics[width=1.1\columnwidth]{figures/icar/frames/frame22.png}
@@
 \end{enumerate}
+Fortran semantics necessitates that gets block.  Puts are non-blocking.
+
 Our results will inform our future decisions around the choice of compiler, communication
@@
 \section{Discussion of Results}
-Results with two different domain sizes (500 x 500 and 2000 x 2000 grid cells) are presented in figure~\ref{fig1-2} for simulations performed on Cheyenne.
-These tests compare scaling with blocking ``get'' and non-blocking ``put'' operations, and compare scaling using \gls{mpi} or OpenSHMEM for communication.
+Figure~\ref{fig1-2} shows results with two domain sizes (500 x 500 and 2000 x 2000 grid cells) running on Cheyenne.
+These tests compare scaling with blocking ``get'' and non-blocking ``put'' operations, and compare scaling using \gls{mpi} or OpenSHMEM.
 The differences between \gls{mpi} and OpenSHMEM are minor, with OpenSHMEM performing only slightly better.
@@
 For the 500x500 domain, ``puts'' scale well to 1800 cores, then scale poorly to \num{3600} cores;
-however, ``gets'' only scale well to \num{504} cores, scale poorly to \num{1224}, an no further improvement comes from up to \num{3600} cores.
-Note \gls{mpi} ``gets'' are not reported because a problem with the \gls{mpi} implementation makes them several orders of magnitude less efficient,
-similarly, \gls{mpi} ``puts'' results are not reported beyond \num{1800} cores because a problem in the \gls{mpi} implementation caused memory allocations to become a bottleneck at higher process counts.
-For the 2000x2000 domain, the same pattern is observed but scaling of ``puts'' remains strong up to 3600 cores;
-scaling of ``gets'' continuous, albeit poorly, to \num{3600} cores.
+however, ``gets'' only scale well to \num{504} cores, scale poorly to \num{1224}, and no further improvement comes from up to \num{3600} cores.
 
-Results across multiple compilers, machines, and architectures are presented in figure~\ref{fig3-4} (top).
+\gls{mpi} ``gets'' are not reported because a problem with the \gls{mpi} implementation makes them several orders of magnitude less efficient.
+Similarly, \gls{mpi} ``put'' results are not reported beyond \num{1800} cores because a problem in the \gls{mpi} implementation caused memory
+allocations to become a bottleneck at higher process counts.
+For the 2000x2000 domain, the same pattern is observed but scaling of ``puts'' remains strong up to 3600 cores; whereas
+scaling of ``gets'' continues, albeit poorly, to \num{3600} cores.
+
+Figure~\ref{fig3-4} (top) depicts results across multiple compilers, machines, and architectures.
 In this case, only ``puts'' are tested, and only for the 2000x2000 domain.
-The GNU compiler was used on Cheyenne (Xeons), and the Cray compiler was used on both Edison (Xeons) and Cori (KNL).
+The GNU compiler was used on Cheyenne (Xeons). The Cray compiler was used on Edison (Xeons) and Cori (KNL).
 To aid interpretation, the scaling results are reported as a fraction of the ideal scaling for each machine in the top left.
@@
 This also shows that the KNL system scales well out to large core counts (60\% ideal with \num{19200} cores),
-but that the total runtimes are significantly slower then the equivalent runtimes on Xeons (top right).
-The KNL performance may be improved in the future by implementing OpenMP threaded parallelism within a node.
+but that the total runtimes are significantly slower than the equivalent runtimes on Xeons (top right).
+The KNL performance might be improved in the future by implementing OpenMP threaded parallelism within a node.
 
 To further test the scaling performance in the best case, ``puts'' on Cray Xeons, a limited set of tests were performed up to nearly \num{100000} cores (fig.~\ref{fig3-4} bottom).
-This system scaled well to over \num{10000} cores, and continued scaling with nearly 50\% of single node efficiency obtained at \num{98304} cores.
+This system scaled well to over \num{10000} cores, and continued scaling with nearly 50\% of the single-node efficiency at \num{98304} cores.
 

### 2017-10-04 body.tex
SWap Chapman citation->OpenSHMEM spec per reviewer
Minor text edits in respose to review

@@
   \includegraphics[width=\textwidth]{figures/fig3_cross_platform.png}
+  \vspace{-6pt}
   \includegraphics[width=\textwidth]{figures/fig4_extreme_scaling.png}
    \caption{Cross-platform and extreme scaling results using \href{https://github.com/gutmann/coarray_icar}{Coarray ICAR}.\label{fig3-4}}
+   \vspace{-6pt}
 \end{figure*}
@@
 \subsection{Objectives}
-The goal of this work is to evaluate several alteranatives for compiling, linking, and executing the
+The goal of this work is to evaluate several alternatives for compiling, linking, and executing the
 same \gls{mini-app} \gls{caf} source code using the following technologies:
 \begin{itemize}
-  \item \gls{mpi}~\cite{mpiforum2016mpi} and OpenSHMEM~\cite{chapman2010introducing} communication layers,
+  \item \gls{mpi}~\cite{mpiforum2016mpi} and OpenSHMEM~\cite{openshmem2016} communication layers,
   \item \gls{gcc} and Cray compilers, and
@@
 We compiled coarray-\gls{icar} on the \gls{nersc} systems using the Cray Fortran compiler version 8.6.0.  We compiled
-at \gls{ncar} using of the \gls{gcc} Fortran (gfortran) compiler front end version 6.3, which uses the
+at \gls{ncar} using the Fortran front end of \gls{gcc} version 6.3, which uses the
 OpenCoarrays \gls{abi}~\cite{fanfarillo2014opencoarrays} for to support \gls{caf}.  We tested two OpenCoarrays parallel runtime libraries:
@@
 
-  This work as funded in part through a contract from the U.S. Army Corps of Engineers.
+  This work was funded in part through a contract from the U.S. Army Corps of Engineers.
 

### 2017-10-05 response.txt
response.txt list edits made for reviewer comments

Attached please find a revised version of our extended abstract accepted for presentation at PAW17.

Reviewer 1
----------
1. Comment: "[2] Please cite the OpenSHMEM 1.3 specification (Available at: http://openshmem.org)"
   Response: The specification is now reference 10.

2. Comment: "Figure 1: The animation is probably not a good idea for the paper. Why not show a few frames instead?"
   Response: Figure 1 now depicts one representative frame. Space does not permit multiple frames.

3. Comment: "Figures 2-: The 'X' marks are indistinguishable when printed. Please use different symbols.
            Please also show error bars rather than plotting over the same point multiple times."
   Response: The symbols have been changed to better differentiate distinct runs.  We did not perform a sufficient
             number of runs to estimate error bars and the different data points are in fact distinct.  For example,
             the close correspondence between the MPI_put and OSH_put points in Figure 2 stems results from separate
             runs in each case and likely relates to commonalities in the underlying hardware or in the MPI and
             OpenSHMEM implementations.

   Comment: "What about the Intel Fortran compiler? Seems like this compiler should be used on KNL, and it should
            support coarrays?"
   Response: Section 1.2 now details our experiences with the Intel compiler, which produces executable programs
             that crash or hang beyond small core counts.

Review 2
--------

1. Comment (a): "The animation in Figure 1 depicts atmospheric processes over North 105 America as simulated by ICAR".
                Which animation? I would suggest take out animation (whatever that is) and put just a figure or snapshots.
   Response: Figure 1 now depicts one representative frame. Space does not permit multiple frames.

2. Comment (b): "Gets: an image that needs data proactively gets the data from memory managed by a providing image
                without the providing image’s active involvement,": what is an image?
                I know what it is, but it is not defined anywhere before this. A reader that knows nothing about CAF
                should at least be able to read what images are.
   Response: The first paragraph in Section 1 now gives a brief overview of the CAF programming model, including the meaning of "image."

3. Comment (c): "Figure 3 (top) depicts results across multiple compilers, machines, and architectures.
                In this case, only “puts” are tested, and only for the 2000x2000 domain.": why only puts? need to justify
                this more.
4. Response: The second paragraph of Section 3 now explains the reasons only "puts" are shown for MPI.

   Comment (d): in figure 3, lower right: title says runtime comparison, but there is no comparison.
   Response: The word "comparison" has now been deleted.

5. Comment (e): line 392 "This work as funded in..." typo.  "This work was funded in"
   Response: This typographical error has been corrected.

Review 3
--------
Comment: "1.2 spelling of "alteranative" needs to be alternative."
Response: The spelling has been corrected.

Comment: "2.2 The sentence 'We compiled at NCAR using of the GCC...." is poorly worded. Restructure the whole sentence
Response: The cited sentence has been resphrased for greater clarity.

### 2017-10-05 body.tex
Complete reviewer-suggested edits

@@
 \subsection{Motivation and Background}
-Since the Fortran 2008 standard was published in 2010~\cite{iso2010information}, a handful of published studies
-have provided encouraging assessments of Fortran's intrinsic parallel programming model, \gls{caf}, in research
-applications running at scale~\cite{preissl2011multithreaded,garain2015comparing,mozdzynski2015partitioned}.   One of the
-most attractive features of \gls{caf} is the ability to express parallel algorithms once within the confines of a single
-language standard without embedding compiler directives or referencing communication layers that are not part of the
-language standard (e.g., no \gls{mpi} or OpenMP).  In theory, this should leave the choice of communication layers
-as a link-time decision.  Ideally, an application developer writes a standard, Fortran-only application, compiles it with any
-standard-conforming compiler, and possibly reserves the option to incorporate any one (or more) of several communication layers
-at link-time.
+Since the Fortran 2008 standard was published in 2010~\cite{iso2010information}, a Fortran program executes as
+a fixed number of instances that the standard refers to as ``images.''  Each image has its own state, including
+data and input/output streams, corresponding to a \gls{spmd} programming style.  Images execute asynchronously
+except at program launch, termination, and program-specified synchronization points.  Images may define
+so-called coarray data structures to provide one-sided access to data by other images in a \gls{pgas}.
+
+A few published studies have provided encouraging assessments of using Fortran's \gls{pgas} features
+in research applications running at scale~\cite{preissl2011multithreaded,garain2015comparing,mozdzynski2015partitioned}.
+One of the most attractive characteristics of these features is the ability to express parallel algorithms in a single
+language without embedding compiler directives or referencing communication layers that are not part of the
+language standard (e.g., no \gls{mpi} or OpenMP).  In theory, this should delay choice of communication layers
+to link-time.  Ideally, an application developer writes a standard, Fortran-only application, compiles it into object code,
+and then links to one or more communication layers.
 
@@
 \gls{icar} simulates the motion of the atmosphere at kilometer length scales and produces flow patterns with a fidelity that is
-attractive to the hydrology community, which studies surface water.  The animation in Figure~\ref{figure:icar} depicts atmospheric
-processes over North America as simulated by \gls{icar}.
+attractive to the hydrology community for studying surface water.  Figure~\ref{figure:icar} depicts results from  \gls{icar}
+simulations over North America.
 
@@
    }
-   \caption{An animation of atmospheric flow of water vapor (blues) and resulting precipitation (green to red) simulated by the \href{https://github.com/gutmann/icar}{\gls{icar}}.
-{\it For the animated version of this graphic, please access https://github.com/sourceryinstitute/coarray-icar-paw17/releases/download/1.0/main.pdf and use Adobe Acrobat Reader 7.0 or later.}
+   \vspace{-24pt}
+   \caption{A visualization of the atmospheric distribution of water vapor (blues) and the resulting precipitation (green to red) simulated by \href{https://github.com/gutmann/icar}{\gls{icar}}.%{\it For the animated version of this graphic, please access https://github.com/sourceryinstitute/coarray-icar-paw17/releases/download/1.0/main.pdf and use Adobe Acrobat Reader 7.0 or later.}
 \label{figure:icar}}
@@
 \subsection{Objectives}
-The goal of this work is to evaluate several alternatives for compiling, linking, and executing the
-same \gls{mini-app} \gls{caf} source code using the following technologies:
+This evaluates alternatives for compiling, linking, and executing one
+\gls{mini-app} \gls{caf} source code using the following technologies:
 \begin{itemize}
@@
 \end{itemize}
-Additionally we show performance differences between two one-sided data access patterns:
+We also tested the Intel Fortran compiler versions 16 and 17. The resulting executable
+program crashes on Cray platforms and hangs at launch on when run on more than 16 cores on the SGI platform.
+
+We show performance for two one-sided data access patterns:
 \begin{enumerate}
-  \item Gets: an image that needs data proactively gets the data from memory managed by a providing
-        image without the providing image's active involvement, and
-  \item Puts: an image that is able to provide data proactively puts the data into memory managed by
-        a receiving image without the receiving image's active involvement.
+  \item Gets: an image retrieves data from memory managed by another
+        image without the providing image's involvement, and
+  \item Puts: an image stores data in memory managed by
+        another image without the receiving image's involvement.
 \end{enumerate}
@@
 
-Our results will inform our future decisions around the choice of compiler, communication
-layer, platform, and data access patterns.
+%Our results will inform our future decisions around the choice of compiler, communication
+%layer, platform, and data access patterns.
 
@@
 
-The test case implemented here uses a classical, idealized hill.
-An idealized 1000 m high mountain is generated in the middle of the domain defined by a sine curve.
-The air is initialized to near 100\% relative humidity, and a background wind of approximately 10 m s$^{-1}$ is specified.
+The test case described here uses an idealized hill: a sine curve representing
+a 1000 m high mountain in the middle of the domain.
+The air is initially near 100\% relative humidity with a background wind of approximately 10 m s$^{-1}$.
 
@@
 This is tested using both a \gls{caf} ``put'' operation, in which one-sided communication is initialized asynchronously to send data to a neighbor,
-and a \gls{caf} ``get'' operation, in which one-sided blocking communication is used to retrieve data from a neighbor.
+and a \gls{caf} ``get'' operation, in which one-sided blocking communication retrieves data from a neighbor.
 
@@
 We performed the experiments presented here on two supercomputers at \gls{nersc},
-located at \gls{lbnl}, and one supercomputer at \gls{ncar}.
-One \gls{lbnl} system is ``Edison,'' a Cray XC30 featuring 5586 nodes with two sockets of 12-core Intel Xeon Processor E5-2695 v2 (``Ivy Bridge''), running at \num{2.4}~\si{\giga\hertz}.
-The second \gls{lbnl} system is ``Cori,'' a Cray XC40 which contains \num{12076} compute nodes, spanning two architectures.
-\num{2388} nodes each have two sockets of 16-core Intel Xeon Processor E5-2698 v3 (``Haswell'') operating at \num{2.3}~\si{\giga\hertz}; the other \num{9688} nodes are single-socket, 68-core Intel Xeon Phi Processor 7250 (``Knights Landing'') at \num{1.4}~\si{\giga\hertz}.
-Both Edison and Cori feature the same ``Aries'' interconnect, with a dragonfly topology.
-All measurements reported here which ran on Cori were executed on the Xeon Phi nodes, running in the ``quadrant`` NUMA configuration, with the MCDRAM configured as a transparent cache to the DDR4 memory.
+located at \gls{lbnl}, and one supercomputer at \gls{ncar}. The \gls{lbnl} systems are
+\begin{itemize}
 
-The \gls{ncar} system is ``Cheyenne,'' an SGI ICE XA Cluster with \num{4032} compute nodes, with two sockets of 18-core Intel Xeon Processor E5-2697 v4 (``Broadwell''), running at \num{2.3}~\si{\giga\hertz}.
-Cheyenne uses a Mellanox EDR infiniband interconnect with a partial 9D Enhanced Hypercube single-plane topology.
+\item Edison: a Cray XC30 featuring 5586 nodes with two sockets of 12-core Intel Xeon Processor E5-2695 v2 (``Ivy Bridge''), running at \num{2.4}~\si{\giga\hertz}.
+\item Cori: a Cray XC40 with \num{12076} compute nodes, spanning two architectures; \num{2388} nodes each have two sockets of 16-core Intel Xeon Processor E5-2698 v3 (``Haswell'') operating at \num{2.3}~\si{\giga\hertz}; the other \num{9688} nodes are single-socket, 68-core Intel Xeon Phi Processor 7250 (``Knights Landing'') at \num{1.4}~\si{\giga\hertz}.
+\end{itemize}
+Edison and Cori have the same ``Aries'' interconnect with a dragonfly topology.
+All Cori measurements on the Xeon Phi nodes ran in the ``quadrant'' NUMA configuration with the MCDRAM configured as a transparent cache to the DDR4 memory.
+
+The \gls{ncar} system is ``Cheyenne,'' an SGI ICE XA Cluster with \num{4032} compute nodes, each containing two sockets of 18-core Intel Xeon Processor E5-2697 v4 (``Broadwell''), running at \num{2.3}~\si{\giga\hertz}.
+Cheyenne uses a Mellanox EDR Infiniband interconnect with a partial 9D Enhanced Hypercube single-plane topology.
 We compiled coarray-\gls{icar} on the \gls{nersc} systems using the Cray Fortran compiler version 8.6.0.  We compiled
-at \gls{ncar} using the Fortran front end of \gls{gcc} version 6.3, which uses the
-OpenCoarrays \gls{abi}~\cite{fanfarillo2014opencoarrays} for to support \gls{caf}.  We tested two OpenCoarrays parallel runtime libraries:
+at \gls{ncar} using the \gls{gcc} version 6.3 Fortran front end, which uses the
+OpenCoarrays \gls{abi}~\cite{fanfarillo2014opencoarrays} to support \gls{caf}.  We tested two OpenCoarrays parallel runtime libraries:
 \begin{enumerate}
   \item The default \gls{mpi} library using the SGI MPT \gls{mpi},
-  \item The openshmem branch using OpenSHMEM.
+  \item The recently released OpenSHMEM library.
 \end{enumerate}
@@
   \item 24 images per Xeon node on Edison, and
-  \item 36 images per node on Cheyenne.
+  \item 36 images per node on Cheyenne,
 \end{itemize}
@@
 \section{Discussion of Results}
-Figure~\ref{fig1-2} shows results with two domain sizes (500 x 500 and 2000 x 2000 grid cells) running on Cheyenne.
+Figure~\ref{fig1-2} shows results with two domain sizes ($500 \times 500 \times 20$ and $2000 \times 2000 \times 20$ grid cells) running on Cheyenne.
 These tests compare scaling with blocking ``get'' and non-blocking ``put'' operations, and compare scaling using \gls{mpi} or OpenSHMEM.
@@
 However, the differences between the ``get'' and ``put'' operations are substantial.
-For the 500x500 domain, ``puts'' scale well to 1800 cores, then scale poorly to \num{3600} cores;
+For the $500 \times 500 \times 20$ domain, ``puts'' scale well to 1800 cores, then scale poorly to \num{3600} cores;
 however, ``gets'' only scale well to \num{504} cores, scale poorly to \num{1224}, and no further improvement comes from up to \num{3600} cores.
@@
 allocations to become a bottleneck at higher process counts.
-For the 2000x2000 domain, the same pattern is observed but scaling of ``puts'' remains strong up to 3600 cores; whereas
+The $2000 \times 2000 \times 20$ domain exhibits the same pattern but the scaling of ``puts'' remains strong up to 3600 cores; whereas
 scaling of ``gets'' continues, albeit poorly, to \num{3600} cores.
@@
 Figure~\ref{fig3-4} (top) depicts results across multiple compilers, machines, and architectures.
-In this case, only ``puts'' are tested, and only for the 2000x2000 domain.
+In this case, only ``puts'' are tested, and only for the $2000 \times 2000\times 20$ domain.
 The GNU compiler was used on Cheyenne (Xeons). The Cray compiler was used on Edison (Xeons) and Cori (KNL).

### 2017-10-05 glossary/acronyms.tex
Add Sec 1.2 Objectives; edit 2.1 Physics &Numerics
Edit Section 2.2 Compilers, runtimes, and hardware
Complete reviewer-suggested edits

@@
+\newacronym{pgas}{PGAS}{partitioned global address space}
+\newacronym{abi}{ABI}{application binary interface}
+\newacronym{gcc}{GCC}{GNU Compiler Collection}
 \newacronym{caf}{CAF}{coarray Fortran}
 \newacronym{mini-app}{mini-app}{mini-application}
-\newacronym{ncar}{NCAR}{National Center for Atmospheric Research}
+\newacronym{ncar}{NCAR}{the National Center for Atmospheric Research}
 \newacronym{wrf-hydro}{WRF-Hydro}{Weather Research and Forcecasting -- Hydrological Model}
-\newacronym{spmd}{SPMD}{Single-Program Multiple-Data}
+\newacronym{spmd}{SPMD}{single-program, multiple-data}
 \newacronym{url}{URL}{Uniform Resource Locator}
@@
 \newacronym{nasa}{NASA}{National Aeronautics and Space Administration}
-\newacronym{nersc}{NERSC}{National Energy Research Scientific Computing Center}
+\newacronym{nersc}{NERSC}{the National Energy Research Scientific Computing Center}
 \newacronym{doe}{DOE}{Department of Energy}

### 2017-10-05 glossary/glossary.tex
Complete reviewer-suggested edits

\newglossaryentry{coarray}{name={coarray},description={a variable that may be referenced or defined by multiple images (replicas) of a Fortran program; the portion of a coarray that is local to a given image may be a scalar or an array}}
\newglossaryentry{refactoring}{name={refactoring},description={changing the decomposition of a code without changing its external behavior}}
\newglossaryentry{image}{name={image},description={instance of a Fortran program}}
\newglossaryentry{exascale}{name={exascale},description={Exascale computing will provide exaFLOPS performance, which refers to $10^{18}$ FLoating-point OPerations per Second (FLOPS)}}
%\newglossaryentry{pgas}{name={PGAS},description={Partitioned Global Address Space, a parallel programming model in which each unit of thread or process or other basic unit of execution contains some portion of a data structure and knows how to access each other execution unit's portion \shortcite{pacheco2011introduction}}}

\newglossaryentry{designpattern}{name={design pattern},description={A common solution to a recurring problem in software design}}
\newglossaryentry{objectorienteddesign}{name={object-oriented design},description={Schematic descriptions of software structure and behavior stressing the relationships between data structures and between data structures and the procedures for operating on the data structures}}
\newglossaryentry{objectorientedprogramming}{name={object-oriented programming},description={A computer program construction philosophy wherein the programmer couples discrete packages of data with a set of procedures for manipulating those data, emphasizing four basic principles: encapsulation, information-hiding, polymorphism, and inheritance}}
\newglossaryentry{softwaredesign}{name={software design},description={A schematic description of the organization and behavior of a computer program}}
\newglossaryentry{encapsulation}{name={encapsulation},description={Bundling data of various intrinsic or programmer-defined types together with procedures for operating on those data.  Fortran supports encapsulation via the definition of derived-type components and type-bound procedure}}
\newglossaryentry{informationhiding}{name={information hiding},description={Limiting access to data or procedures.  Fortran supports information hiding through the ``private'' attribute, the application of which limits the accessibility of data and procedures to code in the same module where the data or procedures are defined}}
\newglossaryentry{class}{name={class},description={An extensible type that encapsulates data along with procedures for operating on those data.  Fortran supports classes via extensible derived types.  An instance of an extensible derived type can be declared with the keyword ``class'' if it is a subprogram dummy argument or if it has either allocatable or pointer attribute}}
\newglossaryentry{derivedtype}{name={derived type},description={A user-defined Fortran type. Fortran user-defined types are by default extensible}}
\newglossaryentry{aggregation} {name={aggregation},description={A ``has a'' or ``whole/part'' relationship between classes wherein an instance of the class representing the ``whole'' encapsulates (has) one or more instances of the class representing the part(s).  Fortran supports aggregation via the components of derived types}}
\newglossaryentry{composition} {name={composition},description={A special case of aggregation wherein the lifetimes (from construction to destruction) of the whole and part coincide}}
\newglossaryentry{inheritance}{name={inheritance},description={An special case of composition which a child class supports procedure invocations as its parent class either by delegating calls to the parent's implementations of those procedures or by overriding the parents' procedures.  Inheritance often referred to as an ``is a'' relationship or as subclassing.  OOP languages support inheritance by automating the insertion of the parent into the child and by automating the procedural delegations.  Fortran supports inheritance via type extension}}
\newglossaryentry{polymorphism}{name={polymorphism},description={The ability for one named type or one named procedure to reference many different types or procedures, respectively, and to resolve any ambiguities in referencing either at compile-time (in the case of \gls{staticpolymorphism}) or at runtime (in the case of \gls{dynamicpolymorphism})}}
\newglossaryentry{staticpolymorphism}{name={static polymorphism},description={A compile-time technology in which the compiler resolves the invocation of a generic procedure name into one of many actual procedure names based on the type, kind, and rank of the procedure's arguments at the point of invocation}}
\newglossaryentry{dynamicpolymorphism}{name={dynamic polymorphism},description={A run-time technology in which the compiler resolves the invocation of a type-bound procedure name into one of many actual procedure names based on the dynamic type of the object on which the procedure is invoked.  The dynamic type may be either the type named in the declaration of the object or any type that extends the named type}}
\newglossaryentry{typeextension}{name={type extension},description={A mechanism for designating that one derived type extends another and thereby inherits the components and type-bound procedures of the latter type}}
\newglossaryentry{instantiate}{name={instantiate},description={To construct an object as an instance of a class.  This typically involves allocating any resources the object requires (e.g., memory) and initializing the object's data.  A mechanism for designating that one derived type extends another and thereby inherits the components and type-bound procedures of the latter type}}
\newglossaryentry{warp}{name={warp},description={A group of threads in a CUDA-enabled device, treated as a unit by the scheduler}}
\newglossaryentry{dummyargument}{name={dummy argument},description={A ``entity whose identifier appears in a dummy argument list in a function or subroutine$\ldots$ or whose name can be used as an argument keyword in a reference to an intrinsic procedure or a procedure in an intrinsic module''~\shortcite{iso2010information}.}}
\newglossaryentry{literalconstant}{name={literal constant},description={A ``constant that does not have a name.''~\shortcite{iso2010information} }}
\newglossaryentry{actualargument}{name={actual argument},description={An ``entity that appears in a procedure reference.''~\shortcite{iso2010information} In the current book, the entity is an expression or a variable}}
\newglossaryentry{defaultinitialization}{name={default initialization},description={A ``mechanism for automatically initializing pointer components to have a defined pointer association status, and nonpointer components to have a particular value .''~\shortcite{iso2010information}.}}
\newglossaryentry{implicitinterface}{name={implicit interface},description={The ``interface of a procedure that includes only the type and type parameters of a function result''~\shortcite{iso2010information}.}}
\newglossaryentry{explicitinterface}{name={explicit interface},description={The ``interface of a procedure that includes all the characteristics of the procedure and names for its dummy arguments except for asterisk dummy arguments.''~\shortcite{iso2010information}.}}
\newglossaryentry{specificinterfaceblock}{name={specific interface block},description={An ``interface block with no generic-spec or ABSTRACT keyword; collection of interface bodies that specify the interfaces of procedures.'' A generic-spec can be a generic name, an operator, or defined input/output generic-spec~\shortcite{iso2010information}.}}
\newglossaryentry{internalsubprogram}{name={internal subprogram},description={A ``subprogram that is contained in a main program or another subprogram.''~\shortcite{iso2010information}.}}
\newglossaryentry{modulesubprogram}{name={module subprogram},description={A ``subprogram that is contained in a module or submodule but is not an internal subprogram.''~\shortcite{iso2010information}.}}
\newglossaryentry{subprogram}{name={subprogram},description={A function or subroutine.~\shortcite{iso2010information}}}
\newglossaryentry{intrinsic}{name={intrinsic},description={Describing a ``type, procedure, module, assignment, operator, or input/output operation defined in [the Fortran standard] and accessible without further definition or specification, or a procedure or module provided by a [compiler] but not defined in'' the standard.''.~\shortcite{iso2010information}.}}
\newglossaryentry{openmp}{name={OpenMP},description={A parallel programming model comprised of compiler directives, library routines, and environment variables define an application program interface for parallelism in C, C++ and Fortran programs. \shortcite{openmp2015}}.}

\newglossaryentry{mpi}{name={MPI},description={The Message Passing Interface, a language-independent communication protocol for programming parallel computers~\shortcite{gropp1999using}}.}
\newglossaryentry{overloading}{name={operator overloading},description={Adding meaning to an intrinsic operator by providing one or more functions for the compiler to invoke when the operator is applied to user-defined derived types.}}

### 2017-10-05 main.tex
Committing edits from August 16.
Delete authordraft documentclass
This step prepares the paper for final submission by
1. Removing watermarks
2. Removig line numbers
Add short title for use in page headers.

@@
-\documentclass[sigconf, authordraft]{acmart}
+\documentclass[sigconf]{acmart}
+%\documentclass[sigconf, authordraft]{acmart}
 
@@
 \begin{document}
-\title{Performance portability of an intermediate-complexity atmospheric research model in coarray Fortran}
+\title[Performance portability of coarray-ICAR]{Performance portability of an intermediate-complexity atmospheric research model in coarray Fortran}
 %\titlenote{Produces the permission block, and
-%  copyright information}
+  %copyright information}
 \subtitle{Extended Abstract}
@@
 results show promising scaling behavior across a range of hardware, compiler, and runtime choices
-on platforms ranging up to 100,000 cores.
+on platforms ranging up to $\sim$100,000 cores.
 \end{abstract}

### 2017-12-11 body.tex
Final submisssion.

@@
 \subsection{Objectives}
-This evaluates alternatives for compiling, linking, and executing one
+This paper evaluates alternatives for compiling, linking, and executing one
 \gls{mini-app} \gls{caf} source code using the following technologies:
@@
 \end{enumerate}
-Fortran semantics necessitates that gets block.  Puts are non-blocking.
-
-%Our results will inform our future decisions around the choice of compiler, communication
-%layer, platform, and data access patterns.
+We also explore the performance and scalability of multi- versus many-core processors.
 
@@
 at \gls{ncar} using the \gls{gcc} version 6.3 Fortran front end, which uses the
-OpenCoarrays \gls{abi}~\cite{fanfarillo2014opencoarrays} to support \gls{caf}.  We tested two OpenCoarrays parallel runtime libraries:
+OpenCoarrays \gls{abi}~\cite{fanfarillo2014opencoarrays} to support \gls{caf}.  We tested two parallel runtime libraries that implement the OpenCoarrays \gls{abi}:
 \begin{enumerate}
@@
 with 75\% efficiency at >10k cores, while on Cheyenne, only 55\% of ideal was achieved.
-This also shows that the KNL system scales well out to large core counts (60\% ideal with \num{19200} cores),
+This also shows that the KNL system scales well out to large core counts (60\% of the ideal with \num{19200} cores),
 but that the total runtimes are significantly slower than the equivalent runtimes on Xeons (top right).

### 2017-12-11 main.tex
Final submisssion.

@@
 a package developed at the \gls{ncar}.
-The \gls{mini-app} uses standard Fortran 2008, including several Fortran 2008 implementations of the collective
-subroutines defined in the Committee Draft of the upcoming Fortran 2015 standard.  The ability of \gls{caf} to run atop various
+The \gls{mini-app} uses standard Fortran 2008, including one Fortran 2008 implementation of a collective
+subroutine defined in the Committee Draft of the upcoming Fortran 2018 standard.  The ability of \gls{caf} to run atop various
 communication layers and the increasing \gls{caf} compiler availability facilitated evaluating several compilers,

### 2017-12-21 main.tex
Add ACM copyright commands
Reformat author list

@@
 \begin{document}
+
+\copyrightyear{2017}
+\acmYear{2017}
+\setcopyright{acmlicensed}
+\acmConference[PAW17]{PAW17: Second Annual PGAS Applications Workshop}{November 12--17, 2017}{Denver, CO, USA}
+\acmPrice{15.00}
+\acmDOI{10.1145/3144779.3169110}
+\acmISBN{978-1-4503-5123-2/17/11}
+
 \title[Performance portability of coarray-ICAR]{Performance portability of an intermediate-complexity atmospheric research model in coarray Fortran}
@@
 
-\author{Damian Rouson}
-\orcid{0000-0002-2234-868X}
-\affiliation{%
-  \institution{Sourcery Institute}
-  \streetaddress{2323 Broadway}
-  \city{Oakland}
-  \state{California}
-  \postcode{94612}
+\author{
+Damian Rouson$^1$, Ethan D Gutmann$^2$, Alessandro Fanfarillo$^2$, Brian Friesen$^3$\vspace{5pt}\\
+\normalsize
+{$^1$Sourcery Institute, USA}\vspace{1pt}\\
+{$^2$National Center for Atmospheric Research, USA}\vspace{1pt}\\
+{$^3$Lawrence Berkeley National Laboratory, USA}\vspace{1pt}\\
 }
-\email{damian@sourceryinstitute.org}
-\renewcommand{\shortauthors}{D. Rouson et al.}
+\renewcommand{\shortauthors}{Damian Rouson et al.}
 
-\author{Ethan D Gutmann}
-\orcid{0000-0003-4077-3430}
-\affiliation{%
-  \institution{National Center for Atmospheric Research}
-  \streetaddress{3450 Mitchell Ln.}
-  \city{Boulder}
-  \state{Colorado}
-  \postcode{80301}
-}
-\email{gutmann@ucar.edu}
+%\author{Damian Rouson}
+%\orcid{0000-0002-2234-868X}
+%\affiliation{%
+  %\institution{Sourcery Institute}
+  %\streetaddress{2323 Broadway}
+  %\city{Oakland}
+  %\state{California}
+  %\postcode{94612}
+%}
+%\email{damian@sourceryinstitute.org}
+%\renewcommand{\shortauthors}{D. Rouson et al.}
+%
+%\author{Ethan D Gutmann}
+%\orcid{0000-0003-4077-3430}
+%\affiliation{%
+  %\institution{National Center for Atmospheric Research}
+  %\streetaddress{3450 Mitchell Ln.}
+  %\city{Boulder}
+  %\state{Colorado}
+  %\postcode{80301}
+%}
+%\email{gutmann@ucar.edu}
+%
+%\author{Brian Friesen}
+%\orcid{0000-0002-1572-1631}
+%\affiliation{%
+  %\institution{Lawrence Berkeley National Laboratory}
+  %\streetaddress{1 Cyclotron Road, Mail Stop 59R4010A}
+  %\city{Berkeley}
+  %\state{California}
+  %\postcode{94720}
+%}
+%\email{BFriesen@lbl.gov}
 
-\author{Brian Friesen}
-\orcid{0000-0002-1572-1631}
-\affiliation{%
-  \institution{Lawrence Berkeley National Laboratory}
-  \streetaddress{1 Cyclotron Road, Mail Stop 59R4010A}
-  \city{Berkeley}
-  \state{California}
-  \postcode{94720}
-}
-\email{BFriesen@lbl.gov}
-
-\author{Alessandro Fanfarillo}
-\orcid{0000-0003-3487-7452}
-\affiliation{%
-  \institution{National Center for Atmospheric Research}
-  \streetaddress{ 1850 Table Mesa Dr.}
-  \city{Boulder}
-  \state{Colorado}
-  \postcode{80305}
-}
-\email{elfanfa@ucar.edu}
+%\author{Alessandro Fanfarillo}
+%\orcid{0000-0003-3487-7452}
+%\affiliation{%
+  %\institution{National Center for Atmospheric Research}
+  %\streetaddress{ 1850 Table Mesa Dr.}
+  %\city{Boulder}
+  %\state{Colorado}
+  %\postcode{80305}
+%}
+%\email{elfanfa@ucar.edu}
 
 \begin{abstract}
-We examine the scalability and performance of an open-source, \gls{caf} \gls{mini-app} that solves
-several parallel, numerical algorithms known to dominate the execution of \gls{icar}~\cite{gutmann2016intermediate} model,
-a package developed at the \gls{ncar}.
-The \gls{mini-app} uses standard Fortran 2008, including one Fortran 2008 implementation of a collective
+We examine the scalability and performance of an open-source, \gls{caf} \gls{mini-app} that implements
+the parallel, numerical algorithms that dominate the execution of \gls{icar}~\cite{gutmann2016intermediate} model
+developed at the \gls{ncar}.
+The Fortran 2008 \gls{mini-app} includes one Fortran 2008 implementation of a collective
 subroutine defined in the Committee Draft of the upcoming Fortran 2018 standard.  The ability of \gls{caf} to run atop various
@@
 different parallel runtime libraries employing one or more communication layers, including \gls{mpi}, OpenSHMEM, and proprietary
-alternatives.  We studied performance on multi- and many-core processors in distributed memory.  The
-results show promising scaling behavior across a range of hardware, compiler, and runtime choices
-on platforms ranging up to $\sim$100,000 cores.
+alternatives.  We study performance on multi- and many-core processors in distributed memory.  The
+results show promising scaling across a range of hardware, compiler, and runtime choices
+on up to $\sim$100,000 cores.
 \end{abstract}
